---
title: "miro"
author:
- name: Helena L. Crowell
  affiliation: CNAG, Barcelona, Spain
output:
  html_document:
    code_folding: show
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
abstract: |
  Todo.
vignette: |
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{miro}
  %\VignetteEncoding{UTF-8}
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('miro')`"
bibliography: "`r file.path(system.file('extdata', package='miro'), 'bib.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center")
library(BiocStyle)
```

# preamble

As a proof of concept, we will process 10x Genomics Visium data comprising
a single tissue section from a colorectal cancer biopsy [@deOliveira2025].

## libs

```{r load-deps}
library(miro)
library(ggplot2)
library(msigdbr)
library(scrapper)
library(VisiumIO)
library(patchwork)
library(OSTA.data)
library(HDF5Array)
library(SpatialExperiment)
library(SpatialExperimentIO)
set.seed(260303)
```

```{r plt}
# utilities for spatial visualization
.df <- \(x) {
    df <- data.frame(colData(x), spatialCoords(x))
    if (length(reducedDims(x))) {
        dr <- as.matrix(reducedDim(x))
        cn <- setdiff(colnames(dr), names(df))
        df <- cbind(df, dr[, cn, drop=FALSE])
    }
    df
}
.xy <- \(z, i, x="x", y="y", theme="void", ...) {
    dots <- list(...)
    if (is.null(dots$size)) dots$size <- 1
    df <- if (!is.data.frame(z)) .df(z) else z
    ok <- length(i) == nrow(df)
    if (ok) { df$foo <- i; i <- "foo" }
    lg <- if (grepl("#", df[[i]][1])) 
        scale_color_identity() else
        switch(scale_type(df[[i]]),
            continuous=scale_color_gradientn(colors=rev(hcl.colors(9, "Batlow"))),
            discrete=scale_color_manual(values=hcl.colors(length(unique(df[[i]])), "Spectral")))
    ggplot(df, aes(.data[[x]], .data[[y]], col=.data[[i]])) +
        do.call(geom_point, dots) + lg +
        coord_equal(expand=FALSE) + 
        get(paste0("theme_", theme))() + 
        theme(
            panel.border=element_rect(),
            plot.title=element_text(hjust=0.5))
}
```

## data

We'll start out by retrieving vendor outputs with `r Biocpkg("OSTA.data")`,
reading data into a `r Biocpkg("SpatialExperiment")` with `r Biocpkg("VisiumIO")`,
and performing minimal processing, namely:

- log-library size normalization
- identifying highly variable features
- principal component analysis (PCA)

```{r load-data}
# retrieval
id <- "Visium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# loading
spe <- TENxVisium(
    spacerangerOut=file.path(td, "outs"),
    images="lowres", format="h5") |>
    import() |> 
    realize()

# wrangling
metadata(spe) <- list()
sym <- rowData(spe)$Symbol
rownames(spe) <- make.unique(sym)
spatialCoordsNames(spe) <- c("x", "y")
img <- as.raster(imgData(spe)$data[[1]])
imgData(spe)$data[[1]] <- SpatialImage(img)
```

Let's visualize the corresponding hematoxylin and eosin (H&E) staining.
On the right-hand side, we crop to the region covered by Visium spots:

```{r plt-he, echo=-1, collapse=TRUE, fig.width=8, fig.height=2.5}
par(mfrow=c(1, 2), mar=c(0,0,0,0))
mx <- as.raster(getImg(spe))
sf <- imgData(spe)$scaleFactor
xy <- sf*spatialCoords(spe)
xy <- ceiling(colRanges(xy))
i <- seq(xy[2, 1], xy[2, 2])
j <- seq(xy[1, 1], xy[1, 2])
plot(mx[rev(seq_len(nrow(mx))), ])
do.call(rect, as.list(c(xy)))
plot(mx[rev(i), j])
```

# analysis

## reduction

```{r pro}
spe <- normalizeRnaCounts.se(spe)
spe <- chooseRnaHvgs.se(spe)
mtx <- logcounts(spe)[rowData(spe)$hvg, ]
    
# principal component analysis (PCA)
(lem <- reducedDim(spe, "PCA") <- calcPca(mtx))
```

## deconvolution

```{r dec}
# retrieve deconvolution results
# (pre-calculated for runtime reasons)
ws <- readRDS(system.file("extdata", "dec.rds", package="miro"))

# scale weights such that they sum to 1
ws <- sweep(ws, 2, colSums(ws), `/`)

# add proportion estimates as metadata
ws <- data.frame(t(as.matrix(ws)))
colData(spe)[names(ws)] <- ws[colnames(spe), ]

# majority vote assignment
mx <- apply(ws, 1, which.max)
spe$RCTD <- factor(names(ws)[mx])

# compartment-level assignment
lab <- list(
    tum="Tumor",
    epi="Intestinal.Epithelial",
    imm=c("B.cells", "T.cells", "Myeloid"),
    str=c("Endothelial", "Fibroblast", "Smooth.Muscle"))
idx <- match(spe$RCTD, unlist(lab))
lab <- rep.int(names(lab), sapply(lab, length))
table(spe$RCTD, spe$sub <- factor(lab[idx]))
```

# visuals

## spatial

Let's start by visualizing the first three PCs over the tissue. In these data,
we can observe that they delineate stroma from healthy and malignant epithelium:

```{r plt-pc, fig.width=10, fig.height=3.5}
cs <- list(
    PC1=c("green", "ivory", "red"), 
    PC2=c("blue",  "ivory", "gold"), 
    PC3=c("black", "grey", "ivory"))
ps <- lapply(paste0("PC", seq_len(3)), \(.) 
    .xy(spe, .) + scale_color_gradientn(colors=cs[[.]]))
wrap_plots(ps, nrow=1) & theme(
      legend.position="bottom",
      legend.text=element_blank())
```

We can now interpret PCs in terms of colors by rescaling and interpreting them
as coordinates in the CIELAB color space (see `?colorspace::LAB()`), where 

- $a\in[-100,100]$ green-red ($<0$ = green, $>0$ = red)
- $b\in[-100,100]$ blue-yellow ($<0$ = blue, $>0$ = yellow) 
- $L\in[0,100]$ lightness or luminance (0 = black, 100 = white).

By default, PCs 1-3 will be used for ABL, but argument `dims`
(default `3L`) can be used to control which components to use.

```{r echo=-1}
options(width=60)
head(df <- calcHex(spe))
add <- setdiff(names(df), names(colData(spe)))
colData(spe) <- cbind(colData(spe), df[add])
```

```{r plt-lab, fig.width=12, fig.height=4}
ps <- list(
    .xy(spe, "hex", "PC1", "PC2", "bw") + 
        ggtitle("principal components (PCs)"),
    .xy(spe, "hex", "A", "B", "bw") + 
        ggtitle("LAB-rescaled PCs"),
    .xy(spe, "hex") + 
        ggtitle("LAB spatially"))
wrap_plots(ps, nrow=1) & scale_color_identity()
```

For comparison, we also visualize deconvolution results spatially:

```{r plt-dec, fig.width=12, fig.height=4}
ps <- list(
    .xy(spe, "RCTD") + ggtitle("deconvolution majority vote"),
    .xy(spe, "sub") + ggtitle("compartment-level label"))
wrap_plots(ps, nrow=1) & 
    theme(legend.key.size=unit(0, "lines")) &
    guides(col=guide_legend(override.aes=list(alpha=1, size=2)))
```

We can also represent higher PCs as LAB, and again visualizing over the tissue
to start seeing variability within domains; e.g., we can observe some variability
within the tumor (left compared to right) and stroma (smooth muscle vs. fibroblasts).

> TODO: legend, tho not sure what that should look like.

```{r plt-xy, message=FALSE, fig.height=4, fig.width=12}
ds <- list(3:5, 5:7, c(3, 5, 7))
ps <- lapply(ds, \(.) {
    lab <- paste("PCs", paste(., collapse=","))
    miroSpatial(spe, col=.) + ggtitle(lab)
})
wrap_plots(ps, nrow=1) & theme(plot.title=element_text(hjust=0.5))
```

## biplot

We can also visualize feature loadings in a biplot with the LAB color space 
as background (note that $L$ remains fixed here). Argument `gs` may be used
to control how many features to include; a couple options are now supported:

- passing a character vector to `gs` or an integer, in which case  
features will be selected according to `how`, where
- `"abs"` gets top-features (per absolute values)
- `"uni"` samples the color space uniformly
- `"min"` gets negatively loaded features
- `"max"` gets positively loaded features

```{r plt-bi, fig.width=10, fig.height=3.5}
ig <- grepv("^IGH", rownames(spe))
miroBiplot(spe, row=10, how="abs") +
miroBiplot(spe, row=20, how="uni") +
miroBiplot(spe, row=ig, L=100, seg=TRUE)
```

## signatures

Because loadings are available for all features underlying PCs
(i.e., the HVGs selected above), we can also query gene sets, e.g.:

```{r msigdb}
# retrieve hallmark gene sets from 'MSigDB'
db <- msigdbr(species="Homo sapiens", collection="H")
# get list of gene symbols, one element per set
gs <- split(db$gene_symbol, db$gs_name)
names(gs) <- gsub("HALLMARK_", "", names(gs))
head(names(gs), 8)
```

We can now project observations into the subspace spanned by these. For this,
`miroSpatial()` supports passing a character vector of genes to project:

```{r plt-sig, fig.width=11, fig.height=7}
gt <- c("GLYCOLYSIS", "APOPTOSIS", "MYC_TARGETS_V2")
ps <- lapply(gt, \(.) {
    map <- calcMap(spe, logcounts(spe), gs[[.]])
    hex <- calcHex(map)$hex
    list(
        .xy(spe, hex) + ggtitle(.),
        miroBiplot(spe, row=gs[[.]]))
})
Reduce(c, ps) |>
    wrap_plots(nrow=2, byrow=FALSE) & 
    theme(plot.title=element_text(hjust=0.5))
```

# interpretation

## loadings

We can now ask something like "Which genes are driving the red region above?", 
which would correspond to positive values for PC1; here, these contain clear
smooth muscle cell markers (MYL9, MHY11, CNN1, DES, TAGLN, ACTG2, etc.):

```{r}
emb <- reducedDim(spe, "PCA")
rot <- featureLoadings(emb)
(gs <- apply(rot[, 1:3], 2, \(.) tail(names(sort(abs(.))), 4)))
```

Let's check the expression of a few of these:

```{r plt-es, fig.width=12, fig.height=9}
lapply(gs, \(.) {
    .xy(spe, logcounts(spe)[., ]) +
    ggtitle(.) }) |> 
    wrap_plots(nrow=3) &
    theme(legend.position="none")
```

We now start seeing a separation of stem-like epithelia (LGR5) towards the left,
with mixed borders reflecting bleeding from surrounding stroma (COL3A1, IGFBP5).

# projection

We'll now load up some Xenium data on a disjoint tissue section from the same publication.

```{r xen}
# retrieve dataset from OSF
id <- "Xenium_HumanColon_Oliveira"
pa <- OSTA.data_load(id, mol=FALSE)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# read into 'SpatialExperiment'
xen <- readXeniumSXE(td, addTx=FALSE)
spatialCoordsNames(xen) <- c("x", "y")
xen <- xen[, xen$transcript_counts > 1]
# area-based normalization
sizeFactors(xen) <- (. <- xen$cell_area)/median(.)
(xen <- normalizeRnaCounts.se(xen))
```

We can now use the Visium-based rotation to compute corresponding PCs.
The LAB representation recapitulates the broad tissue structure from before, 
where red = stroma, blue = healthy, and green = malignant epithelium.

```{r plt-xen, fig.width=12, fig.height=4}
map <- calcMap(spe, logcounts(xen))
hex <- calcHex(map)$hex

# downsample for plotting
idx <- sample(ncol(xen), 1e5)
.xy(xen[, idx], hex[idx], stroke=0, size=1/2)
```

```{r}
# top-loaded
emb <- reducedDim(spe)
rot <- featureLoadings(emb)
(top <- apply(rot[, 1:2], 2, \(.) tail(names(sort(abs(.))), 10)))
```

```{r plt-xen-, fig.width=8, fig.height=5}
# query some interesting genes
row <- unique(as.vector(top))
map <- calcMap(spe, logcounts(xen), row=row)
hex <- calcHex(map)$hex

# downsample for plotting
idx <- sample(ncol(xen), 1e5)
.xy(xen[, idx], hex[idx], stroke=0, size=1/2) + 
miroBiplot(spe, row=row, seg=TRUE) +
plot_layout(widths=c(2, 1))
```

# appendix

<style type="text/css"> .smaller {font-size:10px} </style>

## session {- .smaller}

```{r sess-info, echo=FALSE}
sessionInfo()
```
