---
title: "miro"
author:
- name: Helena L. Crowell
  affiliation: CNAG, Barcelona, Spain

output:
  BiocStyle::html_document:
    code_folding: show
    toc_float: true
    toc_depth: 2
abstract: |
  Todo.
vignette: |
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{miro}
  %\VignetteEncoding{UTF-8}
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('miro')`"
bibliography: "`r file.path(system.file('extdata', package='miro'), 'bib.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center")
library(BiocStyle)
```

# preamble

As a proof of concept, we will process 10x Genomics Visium data comprising
a single tissue section from a colorectal cancer biopsy [@deOliveira2025].

## libs

```{r load-deps}
library(pals)
library(miro)
library(scran)
library(scater)
library(ggplot2)
library(spacexr)
library(msigdbr)
library(VisiumIO)
library(OSTA.data)
library(patchwork)
library(HDF5Array)
library(DropletUtils)
library(SpatialExperiment)
library(SpatialExperimentIO)
set.seed(260303)
```

## data

We'll start out by retrieving vendor outputs with `r Biocpkg("OSTA.data")`,
reading data into a `r Biocpkg("SpatialExperiment")` with `r Biocpkg("VisiumIO")`,
and performing minimal processing, namely:

- log-library size normalization
- identifying highly variable features
- principal component analysis (PCA)

```{r load-data}
# retrieval
id <- "Visium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# loading
spe <- TENxVisium(
    spacerangerOut=file.path(td, "outs"),
    images="lowres", format="h5") |>
    import() |> 
    realize()

# wrangling
metadata(spe) <- list()
sym <- rowData(spe)$Symbol
rownames(spe) <- make.unique(sym)
spatialCoordsNames(spe) <- c("x", "y")
img <- as.raster(imgData(spe)$data[[1]])
imgData(spe)$data[[1]] <- SpatialImage(img)
```

Let's visualize the corresponding hematoxylin and eosin (H&E) staining.
On the right-hand side, we crop to the region covered by Visium spots:

```{r plt-he, echo=-1, fig.width=8, fig.height=2.5}
par(mfrow=c(1, 2), mar=c(0,0,0,0))
mx <- as.raster(getImg(spe))
sf <- imgData(spe)$scaleFactor
xy <- sf*spatialCoords(spe)
xy <- ceiling(colRanges(xy))
i <- seq(xy[2, 1], xy[2, 2])
j <- seq(xy[1, 1], xy[1, 2])
plot(mx[rev(seq_len(nrow(mx))), ])
do.call(rect, as.list(c(xy)))
plot(mx[rev(i), j])
```

# analysis

## reduction

```{r pro}
spe <- logNormCounts(spe)
tbl <- modelGeneVar(spe)
hvg <- getTopHVGs(tbl, n=2e3)
spe <- runPCA(spe, subset_row=hvg)
```

## deconvolution

```{r echo=FALSE}
run <- !file.exists(rds <- "../inst/extdata/dec.rds")
```

```{r sce, eval=run}
# retrieve dataset from OSF repo
id <- "Chromium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# read into 'SingleCellExperiment'
fs <- list.files(td, full.names=TRUE)
h5 <- grep("h5$", fs, value=TRUE)
sce <- read10xCounts(h5, col.names=TRUE)
# add cell metadata
csv <- grep("csv$", fs, value=TRUE)
cd <- read.csv(csv, row.names=1)
colData(sce)[names(cd)] <- cd[colnames(sce), ]
# use gene symbols as feature names
rownames(sce) <- make.unique(rowData(sce)$Symbol)
# exclude cells deemed to be of low-quality
sce <- sce[, sce$QCFilter == "Keep"]
```

```{r dec, eval=run}
# prep reference data (Chromium);
# subset cells from same patient
.sce <- sce[, grepl("P2", sce$Patient)]
# downsample to at most 2,000 cells per cluster
cs <- split(seq_len(ncol(.sce)), .sce$Level1)
cs <- lapply(cs, \(.) sample(., min(length(.), 2e3)))
.sce <- .sce[, unlist(cs)]
# run 'RCTD' deconvolution
rctd_data <- createRctd(spe, .sce, cell_type_col="Level1")
res <- runRctd(rctd_data, max_cores=5, rctd_mode="full")
ws <- assay(res)
```

```{r echo=FALSE}
if (run) saveRDS(ws, rds) else ws <- readRDS(rds)
```

```{r lab}
# scale weights such that they sum to 1
ws <- sweep(ws, 2, colSums(ws), `/`)
# add proportion estimates as metadata
ws <- data.frame(t(as.matrix(ws)))
colData(spe)[names(ws)] <- ws[colnames(spe), ]
# majority vote assignment
mx <- apply(ws, 1, which.max)
spe$RCTD <- factor(names(ws)[mx])
# compartment-level assignment
lab <- list(
    tum="Tumor",
    epi="Intestinal.Epithelial",
    imm=c("B.cells", "T.cells", "Myeloid"),
    str=c("Endothelial", "Fibroblast", "Smooth.Muscle"))
idx <- match(spe$RCTD, unlist(lab))
lab <- rep.int(names(lab), sapply(lab, length))
table(spe$RCTD, spe$Domain <- factor(lab[idx]))
```

# visuals

```{r}
xy <- spatialCoords(spe); pc <- reducedDim(spe, "PCA")
df <- data.frame(fd <- calcHex(spe), colData(spe), xy, pc)
thm <- list(
    theme_bw(), coord_equal(),
    geom_point(size=1, alpha=2/3),
    theme(panel.grid.minor=element_blank()))
col_dis <- scale_color_manual(values=unname(trubetskoy()))
col_con <- scale_color_gradientn(colors=rev(hcl.colors(9, "Batlow")))
```

## spatial

Let's start by visualizing the first three PCs over the tissue. In these data,
we can observe that they delineate stroma from healthy and malignant epithelium:

```{r plt-pc, fig.width=12, fig.height=3}
cs <- list(
    PC1=c("green", "red"), 
    PC2=c("blue", "gold"), 
    PC3=c("black", "grey"))
lapply(paste0("PC", seq_len(3)), 
    \(.) ggplot(df, 
    aes(x, y, col=.data[[.]])) +
    thm + scale_color_gradient2(
        low=cs[[.]][1], high=cs[[.]][2],
        midpoint=sum(range(df[[.]]))/2)) |> 
    wrap_plots(nrow=1)
```

We can now interpret PCs in terms of colors by rescaling and interpreting them
as coordinates in the CIELAB color space (see `?colorspace::LAB()`), where 

- $a\in[-100,100]$ green-red ($<0$ = green, $>0$ = red)
- $b\in[-100,100]$ blue-yellow ($<0$ = blue, $>0$ = yellow) 
- $L\in[0,100]$ lightness or luminance (0 = black, 100 = white).

By default, PCs 1-3 will be used for ABL, but argument `dims`
(default `3L`) can be used to control which components to use.

```{r echo=-1}
options(width=60)
head(fd <- calcHex(spe))
```

```{r plt-lab, fig.width=12, fig.height=4}
ps <- list(
    ggplot(df, aes(PC1, PC2, col=hex)) + ggtitle("principal components (PCs)"),
    ggplot(fd, aes(A, B, col=hex)) + ggtitle("LAB-rescaled PCs"),
    ggplot(df, aes(x, y, col=hex)) + ggtitle("LAB spatially"))
wrap_plots(ps, nrow=1) & thm & scale_color_identity()
```

For comparison, we also visualize deconvolution results spatially:

```{r plt-dec, fig.width=12, fig.height=4}
ps <- list(
    ggplot(df, aes(x, y, col=RCTD)) + ggtitle("deconvolution majority vote"),
    ggplot(df, aes(x, y, col=Domain)) + ggtitle("compartment-level label"))
wrap_plots(ps, nrow=1) & thm & col_dis &
    theme(legend.key.size=unit(0, "lines")) &
    guides(col=guide_legend(override.aes=list(alpha=1)))
```

We can also represent higher PCs as LAB, and again visualizing over the tissue
to start seeing variability within domains; e.g., we can observe some variability
within the tumor (left compared to right) and stroma (smooth muscle vs. fibroblasts).

> TODO: legend, tho not sure what that should look like.

```{r plt-xy, fig.height=4, fig.width=12}
ds <- list(3:5, 5:7, c(3, 5, 7))
ps <- lapply(ds, \(.) miroSpatial(spe, dim=.))
wrap_plots(ps, nrow=1)
```

## biplot

> TODO: not sure if this is right... features good, but colors silghtly off?

We can also visualize feature loadings in a biplot with the LAB color space 
as background (note that $L$ remains fixed here). Argument `gs` may be used
to control how many features to include; a couple options are now supported:

- passing a character vector to `gs` or an integer, in which case  
features will be selected according to `how`, where
- `"abs"` gets top-features (per absolute values)
- `"uni"` samples the color space uniformly
- `"min"` gets negatively loaded features
- `"max"` gets positively loaded features

```{r plt-bi, fig.width=10, fig.height=3}
ig <- grepv("^IGH", rownames(spe))
miroBiplot(spe, gs=10, how="abs") +
miroBiplot(spe, gs=ig, L=100, seg=TRUE) +
miroBiplot(spe, gs=20, how="uni", rad=TRUE)
```

Because loadings are available for all features underlying PCs
(i.e., the HVGs selected above), we can also query gene sets, e.g.:

```{r msigdb, fig.width=16, fig.height=8}
# retrieve hallmark gene sets from 'MSigDB'
db <- msigdbr(species="Homo sapiens", collection="H")
# get list of gene symbols, one element per set
gs <- split(db$gene_symbol, db$gs_name)
names(gs) <- gsub("HALLMARK_", "", names(gs))
# visualize random subset
lapply(names(sample(gs, 8)), \(.) 
    miroBiplot(spe, gs=gs[[.]]) + 
    ggtitle(.)) |> wrap_plots(nrow=2)
```

We can now project observations into the subspace spanned by these. For this,
`miroSpatial()` supports passing a character vector of genes to project:

```{r plt-sig, fig.width=9, fig.height=6}
rot <- attr(pcs <- reducedDim(spe, "PCA"), "rotation")
gt <- c("GLYCOLYSIS", "APOPTOSIS", "MYC_TARGETS_V2")
ps <- lapply(gt, \(.) list(
    miroSpatial(spe, gs=gs[[.]], s=0.4) + ggtitle(.),
    miroBiplot(spe, gs=gs[[.]])))
Reduce(c, ps) |>
    wrap_plots(nrow=2, byrow=FALSE) & 
    theme(plot.title=element_text(hjust=0.5))
```

# interpretation

## loadings

We can now ask something like "Which genes are driving the red region above?", 
which would correspond to positive values for PC1; here, these contain clear
smooth muscle cell markers (MYL9, MHY11, CNN1, DES, TAGLN, ACTG2, etc.):

```{r}
pcs <- reducedDim(spe, "PCA")
rot <- attr(pcs, "rotation")
(gs <- tail(names(sort(rot[, 1])), 14))
```

Let's check the expression of a few of these:

```{r plt-es, fig.width=12, fig.height=6}
ps <- lapply(head(gs, 6), \(.) {
    df[[.]] <- logcounts(spe)[., ]
    ggplot(df, aes(x, y, col=.data[[.]])) 
})
wrap_plots(ps, nrow=2) &
    geom_point(size=1) & coord_equal() &
    theme_bw() & theme(panel.grid.minor=element_blank()) &
    scale_color_gradientn(colors=rev(hcl.colors(9, "Batlow")))
```

## subsetting

> TODO: option to sample color space uniformly, or in a specific direction,
to visualize features of interest for a particular region.

Let's dive deeper into the malignant tissue regions. For simplicity, we here
subset these by filtering on PCs and recomputing PCs on the remaining cells:
(Note that this is a toy example and should not be done in practice.)

```{r sub, fig.width=8, fig.height=4}
idx <- pcs[, 1] < 0 & pcs[, 2] > -20
ncol(sub <- spe[, idx])
tbl <- modelGeneVar(sub)
hvg <- getTopHVGs(tbl, n=2e3)
sub <- runPCA(sub, subset_row=hvg)
miroSpatial(sub) + miroBiplot(sub, seg=TRUE)
```

We now start seeing a separation of stem-like epithelia (LGR5) towards the left,
with mixed borders reflecting bleeding from surrounding stroma (COL3A1, IGFBP5).

# comparison

## UMAP

We'll now compare how UMAP turns out in CIELAB space. Because the orientation 
of xy-axes is arbitrary, colors will change from one run to the next; yet, we'd
expect similar things to get similar colors; surprise, UMAP is misleading...

```{r map}
spe <- runUMAP(spe, dimred="PCA", ncomponents=3, spread=5)
df[colnames(map)] <- (map <- reducedDim(spe, "UMAP"))
```

```{r plt-map, fig.width=15, fig.height=9}
ps <- list(
    p <- miroSpatial(spe) + ggtitle("PCA-based LAB"),
    q <- miroSpatial(spe, "UMAP") + ggtitle("UMAP-based LAB"),
    ggplot(df, aes(x, y, col=RCTD)) + ggtitle("deconvolution") +
        thm + col_dis + theme(legend.key.size=unit(0, "lines")),
    ggplot(df, aes(UMAP1, UMAP2, col=p@data$z)) + 
        thm + scale_color_identity() +
        theme(aspect.ratio=1),
    ggplot(df, aes(UMAP1, UMAP2, col=q@data$z)) + 
        thm + scale_color_identity() +
        theme(aspect.ratio=1),
    ggplot(df, aes(UMAP1, UMAP2, col=RCTD)) + 
        thm + col_dis + theme(
            aspect.ratio=1,
            legend.key.size=unit(0, "lines")))
wrap_plots(ps, nrow=2, guides="collect")
```

## Xenium

We'll now load up some Xenium data on a disjount tissue section from the same publication.

```{r xen}
# retrieve dataset from OSF
id <- "Xenium_HumanColon_Oliveira"
pa <- OSTA.data_load(id, mol=FALSE)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# read into 'SpatialExperiment'
xen <- readXeniumSXE(td, addTx=FALSE)
# area-based normalization
sizeFactors(xen) <- (. <- xen$cell_area)/median(.)
(xen <- logNormCounts(xen))
```

We can now use the Visium-based rotation to compute corresponding PCs.
The LAB representation recapitulates the broad tissue structure from before, 
where red = stroma, blue = healthy, and green = malignant epithelium.

```{r plt-xen, fig.width=8, fig.height=6}
pcs <- reducedDim(spe, "PCA")
rot <- attr(pcs, "rotation")
gt <- intersect(rownames(xen), rownames(rot))
mtx <- t(rot[gt, ]) %*% logcounts(xen)[gt, ]
reducedDim(xen, "PCA") <- as.matrix(t(mtx))
# downsample for plotting
xem <- xen[, sample(ncol(xen), 1e5)]
attr(reducedDim(xem), "rotation") <- rot
miroSpatial(xem, s=0)
```

Let's query some interesting features just because we can:

```{r fig.width=9, fig.height=6}
gt <- grepv("^MUC|KRT|CEACAM|SLC|MMP", rownames(xem))
miroBiplot(xem, gs=gt) +
miroSpatial(xem, s=0, gs=gt) +
plot_layout(widths=c(1, 2))
```

# appendix

<style type="text/css"> .smaller {font-size:10px} </style>

## session {- .smaller}

```{r sess-info, echo=FALSE}
sessionInfo()
```
