---
title: "miro"
author:
- name: Helena L. Crowell
  affiliation: CNAG, Barcelona, Spain
output:
  BiocStyle::html_document:
    code_folding: show
    toc_float: true
    toc_depth: 2
abstract: |
  Todo.
vignette: |
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{miro}
  %\VignetteEncoding{UTF-8}
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('miro')`"
bibliography: "`r file.path(system.file('extdata', package='miro'), 'bib.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center")
library(BiocStyle)
```

# preamble

As a proof of concept, we will process 10x Genomics Visium data comprising
a single tissue section from a colorectal cancer biopsy [@deOliveira2025].

## libs

```{r load-deps}
library(pals)
library(miro)
library(scran)
library(scater)
library(ggplot2)
library(msigdbr)
library(VisiumIO)
library(patchwork)
library(OSTA.data)
library(HDF5Array)
library(SpatialExperiment)
library(SpatialExperimentIO)
set.seed(260303)
```

## data

We'll start out by retrieving vendor outputs with `r Biocpkg("OSTA.data")`,
reading data into a `r Biocpkg("SpatialExperiment")` with `r Biocpkg("VisiumIO")`,
and performing minimal processing, namely:

- log-library size normalization
- identifying highly variable features
- principal component analysis (PCA)

```{r load-data}
# retrieval
id <- "Visium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# loading
spe <- TENxVisium(
    spacerangerOut=file.path(td, "outs"),
    images="lowres", format="h5") |>
    import() |> 
    realize()

# wrangling
metadata(spe) <- list()
sym <- rowData(spe)$Symbol
rownames(spe) <- make.unique(sym)
spatialCoordsNames(spe) <- c("x", "y")
img <- as.raster(imgData(spe)$data[[1]])
imgData(spe)$data[[1]] <- SpatialImage(img)
```

Let's visualize the corresponding hematoxylin and eosin (H&E) staining.
On the right-hand side, we crop to the region covered by Visium spots:

```{r plt-he, echo=-1, fig.width=8, fig.height=2.5}
par(mfrow=c(1, 2), mar=c(0,0,0,0))
mx <- as.raster(getImg(spe))
sf <- imgData(spe)$scaleFactor
xy <- sf*spatialCoords(spe)
xy <- ceiling(colRanges(xy))
i <- seq(xy[2, 1], xy[2, 2])
j <- seq(xy[1, 1], xy[1, 2])
plot(mx[rev(seq_len(nrow(mx))), ])
do.call(rect, as.list(c(xy)))
plot(mx[rev(i), j])
```

# analysis

## reduction

```{r pro}
spe <- logNormCounts(spe)
tbl <- modelGeneVar(spe)
hvg <- getTopHVGs(tbl, n=2e3)
spe <- runPCA(spe, subset_row=hvg)
```

## deconvolution

```{r dec}
# retrieve deconvolution results
# (pre-calculated for runtime reasons)
ws <- readRDS(system.file("extdata", "dec.rds", package="miro"))

# scale weights such that they sum to 1
ws <- sweep(ws, 2, colSums(ws), `/`)

# add proportion estimates as metadata
ws <- data.frame(t(as.matrix(ws)))
colData(spe)[names(ws)] <- ws[colnames(spe), ]

# majority vote assignment
mx <- apply(ws, 1, which.max)
spe$RCTD <- factor(names(ws)[mx])

# compartment-level assignment
lab <- list(
    tum="Tumor",
    epi="Intestinal.Epithelial",
    imm=c("B.cells", "T.cells", "Myeloid"),
    str=c("Endothelial", "Fibroblast", "Smooth.Muscle"))
idx <- match(spe$RCTD, unlist(lab))
lab <- rep.int(names(lab), sapply(lab, length))
table(spe$RCTD, spe$Domain <- factor(lab[idx]))
```

# visuals

```{r vis}
xy <- spatialCoords(spe); pc <- reducedDim(spe, "PCA")
df <- data.frame(fd <- calcHex(spe), colData(spe), xy, pc)
thm <- \(.="void") list(
    get(paste0("theme_", .))(),
    geom_point(size=1, alpha=2/3),
    coord_equal(), theme(
        panel.border=element_rect(),
        plot.title=element_text(hjust=0.5)))
col_dis <- scale_color_manual(values=unname(trubetskoy()))
col_con <- scale_color_gradientn(colors=rev(hcl.colors(9, "Batlow")))
```

## spatial

Let's start by visualizing the first three PCs over the tissue. In these data,
we can observe that they delineate stroma from healthy and malignant epithelium:

```{r plt-pc, fig.width=9, fig.height=3.5}
cs <- list(
    PC1=c("green", "white", "red"), 
    PC2=c("blue",  "white", "gold"), 
    PC3=c("black", "white", "grey"))
ps <- lapply(paste0("PC", seq_len(3)), \(.) {
    pal <- scale_color_gradientn(colors=cs[[.]])
    ggplot(df, aes(x, y, col=.data[[.]])) + pal + thm()
})
wrap_plots(ps, nrow=1) & theme(
      legend.position="bottom",
      legend.text=element_blank())
```

We can now interpret PCs in terms of colors by rescaling and interpreting them
as coordinates in the CIELAB color space (see `?colorspace::LAB()`), where 

- $a\in[-100,100]$ green-red ($<0$ = green, $>0$ = red)
- $b\in[-100,100]$ blue-yellow ($<0$ = blue, $>0$ = yellow) 
- $L\in[0,100]$ lightness or luminance (0 = black, 100 = white).

By default, PCs 1-3 will be used for ABL, but argument `dims`
(default `3L`) can be used to control which components to use.

```{r echo=-1}
options(width=60)
head(fd <- calcHex(spe))
```

```{r plt-lab, fig.width=12, fig.height=4}
ps <- list(
    ggplot(df, aes(PC1, PC2, col=hex)) + 
        thm("bw") + ggtitle("principal components (PCs)"),
    ggplot(fd, aes(A, B, col=hex)) + 
        thm("bw") + ggtitle("LAB-rescaled PCs"),
    ggplot(df, aes(x, y, col=hex)) + 
        thm() + ggtitle("LAB spatially"))
wrap_plots(ps, nrow=1) & scale_color_identity()
```

For comparison, we also visualize deconvolution results spatially:

```{r plt-dec, fig.width=12, fig.height=4}
ps <- list(
    ggplot(df, aes(x, y, col=RCTD)) + thm() + ggtitle("deconvolution majority vote"),
    ggplot(df, aes(x, y, col=Domain)) + thm() + ggtitle("compartment-level label"))
wrap_plots(ps, nrow=1) & col_dis &
    theme(legend.key.size=unit(0, "lines")) &
    guides(col=guide_legend(override.aes=list(alpha=1, size=2)))
```

We can also represent higher PCs as LAB, and again visualizing over the tissue
to start seeing variability within domains; e.g., we can observe some variability
within the tumor (left compared to right) and stroma (smooth muscle vs. fibroblasts).

> TODO: legend, tho not sure what that should look like.

```{r plt-xy, message=FALSE, fig.height=4, fig.width=12}
ds <- list(3:5, 5:7, c(3, 5, 7))
ps <- lapply(ds, \(.) {
    lab <- paste("PCs", paste(., collapse=","))
    miroSpatial(spe, dim=.) + ggtitle(lab)
})
wrap_plots(ps, nrow=1) & theme(plot.title=element_text(hjust=0.5))
```

## biplot

> TODO: not sure if this is right... features good, but colors silghtly off?

We can also visualize feature loadings in a biplot with the LAB color space 
as background (note that $L$ remains fixed here). Argument `gs` may be used
to control how many features to include; a couple options are now supported:

- passing a character vector to `gs` or an integer, in which case  
features will be selected according to `how`, where
- `"abs"` gets top-features (per absolute values)
- `"uni"` samples the color space uniformly
- `"min"` gets negatively loaded features
- `"max"` gets positively loaded features

```{r plt-bi, fig.width=10, fig.height=3.5}
ig <- grepv("^IGH", rownames(spe))
miroBiplot(spe, gs=10, how="abs") +
miroBiplot(spe, gs=20, how="uni") +
miroBiplot(spe, gs=ig, L=100, seg=TRUE)
```

## signatures

Because loadings are available for all features underlying PCs
(i.e., the HVGs selected above), we can also query gene sets, e.g.:

```{r msigdb}
# retrieve hallmark gene sets from 'MSigDB'
db <- msigdbr(species="Homo sapiens", collection="H")
# get list of gene symbols, one element per set
gs <- split(db$gene_symbol, db$gs_name)
names(gs) <- gsub("HALLMARK_", "", names(gs))
head(names(gs), 8)
```

We can now project observations into the subspace spanned by these. For this,
`miroSpatial()` supports passing a character vector of genes to project:

```{r plt-sig, fig.width=10, fig.height=6}
gt <- c("GLYCOLYSIS", "APOPTOSIS", "MYC_TARGETS_V2")
ps <- lapply(gt, \(.) list(
    miroSpatial(spe, gs=gs[[.]], size=1/2) + ggtitle(.),
    miroBiplot(spe, gs=gs[[.]])))
Reduce(c, ps) |>
    wrap_plots(nrow=2, byrow=FALSE) & 
    theme(plot.title=element_text(hjust=0.5))
```

## regression

```{r pcr, echo=-1, fig.width=8, fig.height=4}
par(mar=c(4,4,2,0))
xy <- spatialCoords(spe)
pc <- reducedDim(spe, "PCA")
lm <- summary(lm(pc ~ xy))
r2 <- sapply(lm, \(.) .$r.squared)
plot(r2, type="o", 
    xlab="principal component (PC)", ylab="R-squared",
    main="regression of spatial coordinates against PCs")
```

```{r fig.width=9, fig.height=7}
ds <- list(
    # not interesting
    c(6,8,10), 16:18, 22:24,
    # interesting?
    c(4,7,9), c(11,12,15), 25:27)
ps <- lapply(ds, \(.) {
    lab <- paste("PCs", paste(., collapse=","))
    miroSpatial(spe, dim=., size=2/3) + ggtitle(lab)
})
wrap_plots(ps, nrow=2) &
theme(plot.title=element_text(hjust=0.5))
```

```{r}
# checking out PC8, which has low R2;
# looks like EC and PC markers...
tail(names(sort(abs(attr(pc, "rotation")[, 8]))), 20)
```

# interpretation

## loadings

We can now ask something like "Which genes are driving the red region above?", 
which would correspond to positive values for PC1; here, these contain clear
smooth muscle cell markers (MYL9, MHY11, CNN1, DES, TAGLN, ACTG2, etc.):

```{r}
pcs <- reducedDim(spe, "PCA")
rot <- attr(pcs, "rotation")
(gs <- tail(names(sort(rot[, 1])), 14))
```

Let's check the expression of a few of these:

```{r plt-es, fig.width=12, fig.height=6}
ps <- lapply(head(gs, 6), \(.) {
    df[[.]] <- logcounts(spe)[., ]
    ggplot(df, aes(x, y, col=.data[[.]])) 
})
wrap_plots(ps, nrow=2) &
    geom_point(size=1/2) & coord_equal() & thm() &
    scale_color_gradientn(colors=rev(hcl.colors(9, "Batlow")))
```

## subsetting

> TODO: option to sample color space in a specific direction.

Let's dive deeper into the malignant tissue regions. For simplicity, we here
subset these by filtering on PCs and recomputing PCs on the remaining cells:
(Note that this is a toy example and should not be done in practice.)

```{r sub, fig.width=8, fig.height=4}
idx <- pcs[, 1] < 0 & pcs[, 2] > -20
ncol(sub <- spe[, idx])
tbl <- modelGeneVar(sub)
hvg <- getTopHVGs(tbl, n=2e3)
sub <- runPCA(sub, subset_row=hvg)
miroSpatial(sub) + miroBiplot(sub, gs=33, seg=TRUE)
```

We now start seeing a separation of stem-like epithelia (LGR5) towards the left,
with mixed borders reflecting bleeding from surrounding stroma (COL3A1, IGFBP5).

# projection

We'll now load up some Xenium data on a disjoint tissue section from the same publication.

```{r xen}
# retrieve dataset from OSF
id <- "Xenium_HumanColon_Oliveira"
pa <- OSTA.data_load(id, mol=FALSE)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
# read into 'SpatialExperiment'
xen <- readXeniumSXE(td, addTx=FALSE)
# area-based normalization
sizeFactors(xen) <- (. <- xen$cell_area)/median(.)
(xen <- logNormCounts(xen))
```

We can now use the Visium-based rotation to compute corresponding PCs.
The LAB representation recapitulates the broad tissue structure from before, 
where red = stroma, blue = healthy, and green = malignant epithelium.

```{r plt-xen, fig.width=12, fig.height=4}
pcs <- reducedDim(spe, "PCA")
rot <- attr(pcs, "rotation")
gt <- intersect(rownames(xen), rownames(rot))
mtx <- t(rot[gt, ]) %*% logcounts(xen)[gt, ]
reducedDim(xen, "PCA") <- as.matrix(t(mtx))

# downsample for plotting
xem <- xen[, sample(ncol(xen), 1e5)]
attr(reducedDim(xem), "rotation") <- rot
miroSpatial(xem, size=0, shape=16)
```

Let's query some highly loaded features just because we can:

```{r fig.width=9, fig.height=6}
p <- miroBiplot(xem, gs=33, how="abs")
gs <- get_layer_data(p , 5)$label
p + miroSpatial(xem, size=0, gs=gs) + plot_layout(widths=c(1, 2))
```

# appendix

<style type="text/css"> .smaller {font-size:10px} </style>

## session {- .smaller}

```{r sess-info, echo=FALSE}
sessionInfo()
```
