---
title: "miro"
author:
- name: Helena L. Crowell
  affiliation: CNAG, Barcelona, Spain

output:
  BiocStyle::html_document:
    code_folding: show
abstract: |
  Todo.
vignette: |
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{miro}
  %\VignetteEncoding{UTF-8}
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('miro')`"
bibliography: "`r file.path(system.file('extdata', package='miro'), 'bib.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center")
library(BiocStyle)
```

# preamble

As a proof of concept, we will process 10x Genomics Visium data comprising
a single tissue section from a colorectal cancer biopsy [@deOliveira2025].

## libs

```{r load-deps}
library(miro)
library(scran)
library(scater)
library(ggplot2)
library(VisiumIO)
library(OSTA.data)
library(patchwork)
library(HDF5Array)
library(SpatialExperiment)
set.seed(260303)
```

## data

We'll start out by retrieving vendor outputs with `r Biocpkg("OSTA.data")``,
reading data into a `r Biocpkg("SpatialExperiment")` with `r Biocpkg("VisiumIO")`,
and performing minimal processing, namely:
- log-library size normalization
- identifying highly variable features
- principal component analysis (PCA)

```{r load-data, cache=TRUE}
# retrieval
id <- "Visium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# loading
spe <- TENxVisium(
    spacerangerOut=file.path(td, "outs"),
    images="lowres", format="h5") |>
    import() |> 
    realize()

# wrangling
metadata(spe) <- list()
rownames(spe) <- rowData(spe)$Symbol
spatialCoordsNames(spe) <- c("x", "y")
img <- as.raster(imgData(spe)$data[[1]])
imgData(spe)$data[[1]] <- SpatialImage(img)

# processing
spe <- logNormCounts(spe)
tbl <- modelGeneVar(spe)
hvg <- getTopHVGs(tbl, n=2e3)
spe <- runPCA(spe, subset_row=hvg)
spe
```

Let's visualize the corresponding hematoxylin and eosin (H&E) staining.
On the right-hand side, we crop to the region covered by Visium spots:

```{r plt-he, echo=-1, fig.width=8, fig.height=2.5}
par(mfrow=c(1, 2), mar=c(0,0,0,0))
mx <- as.raster(getImg(spe))
sf <- imgData(spe)$scaleFactor
xy <- sf*spatialCoords(spe)
xy <- ceiling(colRanges(xy))
i <- seq(xy[2, 1], xy[2, 2])
j <- seq(xy[1, 1], xy[1, 2])
plot(mx[rev(seq_len(nrow(mx))), ])
do.call(rect, as.list(c(xy)))
plot(mx[rev(i), j])
```

# analysis

We can now interpret PCs in terms of colors by rescaling and interpreting them
as coordinates in the CIELAB color space (see `?colorspace::LAB()`), where 

- $a\in[-100,100]$ green-red ($<0$ = green, $>0$ = red)
- $b\in[-100,100]$ blue-yellow ($<0$ = blue, $>0$ = yellow) 
- $L\in[0,100]$ lightness or luminance (0 = black, 100 = white).

By default, PCs 1-3 will be used for ABL, but argument `dims`
(default `3L`) can be used to control which components to use.

```{r echo=-1}
options(width=60)
head(df <- calcHex(spe))
```

```{r fig.width=12, fig.height=6, class.source="fold-hide"}
pc <- reducedDim(spe, "PCA")
ts <- c(0, -0.5*pi, -pi, -1.5*pi)
names(ts) <- c("0", "-pi/2", "-pi", "-3pi/2")
lapply(names(ts), \(.) {
    fd <- calcHex(spe, t=ts[.])
    df <- data.frame(pc, fd)
    ggplot(fd, aes(A, B, col=hex)) + 
    ggtitle(paste("theta =", .)) +
    ggplot(df, aes(PC1, PC2, col=hex)) +
    plot_layout(ncol=1) &
        geom_point(alpha=0.5) &
        scale_color_identity() &
        theme_bw() & coord_equal() &
        theme(panel.grid.minor=element_blank())
}) |> wrap_plots(nrow=1)
```

# visuals

## spatial

> TODO: legend, tho not sure what that should look like.

```{r plt-xy, fig.height=4, fig.width=12}
ds <- list(1:3, 3:5, 5:7)
ps <- lapply(ds, \(.) miroSpatial(spe, dim=.))
wrap_plots(ps, nrow=1)
```

## biplot

> TODO: come up w better viz, e.g., circular with repeling labels?

We can also visualize feature loadings in a biplot with the LAB color space 
as background (note that $L$ remains fixed here). Argument `gs` may be used
to control which features to include; this can be either a character vector 
or scalar integer, in which case the top-`gs` features in each dimension 
will be determined (per absolute values).

```{r plt-bi, fig.width=6, fig.height=3}
ig <- grepv("^IGH", rownames(spe))
miroBiplot(spe, gs=5L) +
miroBiplot(spe, gs=ig, L=100)
```

# interpretation

## loadings

We can now ask something like "Which genes are driving the red region above?", 
which would correspond to positive values for PC1; here, these contain clear
smooth muscle cell markers (MYL9, MHY11, CNN1, DES, TAGLN, ACTG2, etc.):

```{r}
pcs <- reducedDim(spe, "PCA")
rot <- attr(pcs, "rotation")
(gs <- tail(names(sort(rot[, 1])), 14))
```

Let's check the expression of a few of these:

```{r plt-es, fig.width=12, fig.height=6}
df <- data.frame(spatialCoords(spe))
ps <- lapply(head(gs, 6), \(.) {
    df[[.]] <- logcounts(spe)[., ]
    ggplot(df, aes(x, y, col=.data[[.]])) 
})
wrap_plots(ps, nrow=2) &
    geom_point(size=1) & coord_equal() &
    theme_bw() & theme(panel.grid.minor=element_blank()) &
    scale_color_gradientn(colors=rev(hcl.colors(9, "Batlow")))
```

## subsetting

Let's dive deeper into the malignant tissue regions. For simplicity, we here
subset these by filtering on PCs and recomputing PCs on the remaining cells:
(Note that this is a toy example and should not be done in practice.)

```{r sub, fig.width=8, fig.height=4}
idx <- pcs[, 1] < 0 & pcs[, 2] > -20
ncol(sub <- spe[, idx])
tbl <- modelGeneVar(sub)
hvg <- getTopHVGs(tbl, n=2e3)
sub <- runPCA(sub, subset_row=hvg)
miroSpatial(sub) + miroBiplot(sub)
```

We now start seeing a separation of stem-like epithelia (LGR5) towards the left,
with mixed borders reflecting bleeding from surrounding stroma (COL3A1, IGFBP5).

# appendix

<style type="text/css"> .smaller {font-size:10px} </style>

## session {- .smaller}

```{r sess-info, echo=FALSE}
sessionInfo()
```
