---
title: "miro"
author:
- name: Helena L. Crowell
  affiliation: CNAG, Barcelona, Spain

output:
  BiocStyle::html_document:
    code_folding: show
abstract: |
  Todo.
vignette: |
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{miro}
  %\VignetteEncoding{UTF-8}
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('miro')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.align="center")
```

# preamble

## libs

```{r load-deps}
library(miro)
library(scran)
library(scater)
library(ggplot2)
library(VisiumIO)
library(OSTA.data)
library(patchwork)
library(HDF5Array)
set.seed(260303)
```

## data

```{r load-data, cache=TRUE}
# retrieval
id <- "Visium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# loading
spe <- TENxVisium(
    spacerangerOut=file.path(td, "outs"),
    images="lowres", format="h5") |>
    import()
rownames(spe) <- rowData(spe)$Symbol
metadata(spe) <- list()

# processing
spe <- logNormCounts(spe)
tbl <- modelGeneVar(spe)
hvg <- getTopHVGs(tbl, n=2e3)
spe <- runPCA(spe, subset_row=hvg)
spe
```

# analysis

```{r echo=-1}
options(width=60)
head(df <- calcHex(spe))
```

```{r fig.width=12, fig.height=6, class.source="fold-hide"}
pc <- reducedDim(spe, "PCA")
ts <- c(0, -0.5*pi, -pi, -1.5*pi)
names(ts) <- c("0", "-pi/2", "-pi", "-3pi/2")
lapply(names(ts), \(.) {
    fd <- calcHex(spe, t=ts[.])
    df <- data.frame(pc, fd)
    ggplot(fd, aes(A, B, col=hex)) + 
    ggtitle(paste("theta =", .)) +
    ggplot(df, aes(PC1, PC2, col=hex)) +
    plot_layout(ncol=1) &
        geom_point(alpha=0.5) &
        scale_color_identity() &
        theme_bw() & coord_equal() &
        theme(panel.grid.minor=element_blank())
}) |> wrap_plots(nrow=1)
```

# visuals

## spatial

> TODO: legend, tho not sure what that should look like.

```{r plt-xy, fig.height=4, fig.width=12}
ds <- list(1:3, 3:5, 5:7)
ps <- lapply(ds, \(.) miroSpatial(spe, dim=.))
wrap_plots(ps, nrow=1)
```

## biplot

> TODO: come up w better viz, e.g., circular with repeling labels?

- `dims`: dimensions to use
- `L`: luminance; passed to `LAB()`
- `gs`: features to use; can be a character vector or scalar integer,  
  in which case the top-`gs` features per dimension will be used

```{r plt-bi, fig.width=6, fig.height=3}
ig <- grepv("^IGH", rownames(spe))
miroBiplot(spe, gs=5L) +
miroBiplot(spe, gs=ig, L=100)
```

# appendix

<style type="text/css"> .smaller {font-size:10px} </style>

## session {- .smaller}

```{r sess-info, echo=FALSE}
sessionInfo()
```
